Dynamical modeling of tidal streams

### 2013-12-24 tex/ms.tex
streams: intro and re-worded abstract

\begin{abstract} 
  I present a new framework for modeling the dynamics of cold tidal
  streams that is useful for constraining the gravitational host
  potential and progenitor properties from observed tidal-stream data
  and for quickly generating mock tidal streams. The framework
  consists of simple models for the initial action--angle distribution
  of tidal debris as a function of time, which can be
  straightforwardly evolved forward in time. Taking advantage of the
  essentially one-dimensional nature of cold tidal streams, the
  transformation to position--velocity coordinates can be linearized
  and interpolated near a small number of points along the progenitor
  orbit, thus allowing for efficient computations of a stream's
  properties in observable quantities. I illustrate how to calculate
  the stream's track in different coordinate systems, how to quickly
  estimate the dispersion perpendicular to the track, and how to draw
  mock stream data. As a generative model, this framework allows one
  to compute the full probability distribution function and
  marginalize over or condition it on certain phase--space dimensions
  as well as convolve it with observational uncertainties. This will
  be instrumental in proper data analysis of stream data. In addition
  to providing a computationally-efficient practical tool for modeling
  the dynamics of tidal streams, the action--angle nature of the
  framework helps elucidate exactly in what manner streams do not
  follow single orbits, how the observed width of the stream relates
  to the velocity dispersion or mass of the progenitor, and how the
  progenitor of ``orphan'' streams could be located.

@@

@@

The dynamics of tidal streams both in real (position--velocity) and
action--angle space has been described by various previous authors
\citep[\eg,][]{Helmi99a,Tremaine99a,SomeJohnstonPaper,Sanders13a}. I
provide here a brief description of the dynamics of tidal streams that
is relevant for what follows.

@@

Thus, in action--angle coordinates, the dynamics of tidal-stream
formation is simplified to linear growth of small initial angle
differences due to small action/frequency differences. The angle
direction $\unitvec{\veco} \equiv \Delta \veco / | \Delta \veco|$
along which a star moves away from the progenitor is related to the
initial action offset by $\Delta \veco \approx (\partial^2 H
/ \partial \vecj \partial \vecj) \Delta \vecj$, in the limit of small
action offsets, as $\veco = \partial H / \partial \vecj$. For an
approximately one-dimensional stream to form, the Hessian matrix
$(\partial^2 H / \partial \vecj \partial \vecj)\large|_{\vecj^p}$ has
to be dominated by a single large eigenvalue; in this case the angle
difference of stars stripped from the stream at any time will fall
along approximately the same direction $\unitvec\veco$, and the stream
will essentially be one dimensional. In detail, because the
distribution of initial offsets $\Delta \vecj$ is not isotropic, the
direction $\unitvec \veco$ is not just determined by the potential
(through $H$) and the progenitor orbit, but also by the distribution
of $\Delta \vecj$, which is set through a combination of the internal
dynamics of the progenitor and the stripping process (see discussion
in \citealt{Sanders13a}). For dynamical modeling we can be agnostic
about the relative importance of all of these contributors to
$\unitvec \veco$ by flexible modeling of the distribution of $\Delta
\vecj$.

@@

The probability in \eqnname~(\ref{eq:aadfsimple}) incorporates and
elucidates commonly used methods for fitting tidal streams. On the one
hand are approaches that search to minimize the spread in energy,
integrals of the motion, or actions (\eg,
\citealt{Binney08a,Penarrubia12a}, Sanderson \etal, 2013, in
preparation). These approaches ignore the correlations between actions
and angles in the stream (the exponential in
\eqnname~(\ref{eq:aadfsimple}), which describes this correlation
through the frequencies) and only use $f_{\vecj}(\Delta
\vecj_i|\paramsdiff)$; finding the gravitational potential that
minimizes the spread in actions optimizes $f_{\vecj}(\Delta
\vecj_i|\paramsdiff)$. Recently, a method was proposed that only uses
the correlation between the actions and the angles, expressed by the
exponential in \eqnname~(\ref{eq:aadfsimple}) \citep{Sanders13a}; this
method does not use the fact that the spread in actions in a tidal
stream is small (expressed by $f_{\vecj}(\Delta
\vecj_i|\paramsdiff)$).

@@

\bibitem[{{Binney} \& {Tremaine}(2008)}]{binneytremaine}
  Binney,~J. \& Tremaine,~S. 2008, Galactic Dynamics: Second Edition
\bibitem[Binney(2008)]{Binney08a}
  Binney,~J. 2008, \mnras, 386, L47
\bibitem[Eyre \& Binney(2011)]{Eyre11a}
  Eyre,~A. \& Binney,~J. 2011, \mnras, 413, 1852
\bibitem[Pe{\~n}arrubia \etal(2012)]{Penarrubia12a}
  Pe{\~n}arrubia, J., Koposov,~S.~E., \& Walker,~M.~G. 2012, \apj, 760, 2
\bibitem[Sanders \& Binney(2013)]{Sanders13a}
  Sanders,~J. \& Binney,~J. 2013, \mnras, 433, 1826
\end{thebibliography}

>>>>

\begin{abstract} 
  I present a new framework for modeling the dynamics of cold tidal
  streams. The framework consists of simple models for the initial
  action--angle distribution of tidal debris as a function of time,
  which can be straightforwardly evolved forward in time. Taking
  advantage of the essentially one-dimensional nature of cold tidal
  streams, the transformation to position--velocity coordinates can be
  linearized and interpolated near a small number of points along the
  progenitor orbit, thus allowing for efficient computations of a
  stream's properties in observable quantities. I illustrate how to
  calculate the stream's track in different coordinate systems, how to
  quickly estimate the dispersion perpendicular to the track, and how
  to draw mock stream data. As a generative model, this framework
  allows one to compute the full probability distribution function and
  marginalize over or condition it on certain phase--space dimensions
  as well as convolve it with observational uncertainties. This will
  be instrumental in proper data analysis of stream data. In addition
  to providing a computationally-efficient practical tool for modeling
  the dynamics of tidal streams, the action--angle nature of the
  framework helps elucidate in exactly what manner streams do not
  follow single orbits, how the observed width of the stream relates
  to the velocity dispersion or mass of the progenitor, and how the
  progenitor of ``orphan'' streams could be located.

@@

Tidal streams hold enormous promise as probes of both the large-scale
structure of the Milky Way halo's density distribution
\citep[\eg,][]{Johnston99a,Koposov10a} and its small-scale
fluctuations \citep{Carlberg09a,Carlberg12a}. However, two factors
have hampered the practical use of tidal streams as constraints on the
gravitational potential. First, streams do not trace a single orbit,
which has led to confusion over how problematic the single-orbit
approximation really is \citep{Eyre11a,Sanders13a}. Second, approaches
that go beyond the single-orbit assumption encounter practical
difficulties of computational cost and observational data quality
\citep{Sanders13b,PriceWhelan13a}. Similarly, spurious underdensities
created by the dynamics of stream stars hinder our ability to use
underdensities in star counts along a tidal stream in order to
constrain the number of encounters with dark satellites and their
masses \citep{Ngan14a}. Fast, generative models of tidal streams would
greatly benefit both of these applications.

It has long been clear that the dynamics of tidal streams is most
simply described in terms of action--angle coordinates
\citep{Tremaine99a,Helmi99a}. Once a star has been tidally stripped
from the progenitor, the self-gravity of the stream can be neglected
and the orbital actions of a stream member are conserved while the
angles increase linearly with time. The action--angle structure of a
stream is therefore characterized by strong but straightforward
correlations between the actions and angles of stream members, the
exploitation of which is crucial for using streams to measure the host
gravitational potential \citep{Sanders13b}. The description of a
stream in action--angle coordinates also makes the connection between
the orbit and velocity dispersion (or mass) of the progenitor and the
action distribution of the tidal debris clear \citep{Eyre11a},
allowing simple physical models of the stream to be specified.

However, streams are observed in position--velocity coordinates and
for action--angle descriptions to be useful, we must be able to
calculate the transformation between these coordinate systems
efficiently. Until now this has required accurate phase--space data
\citep{Sanders13b} and specialized algorithms to calculate
action--angle coordinates that break down for the eccentric orbits on
which tidal-stream progenitors are typically found
\citep{Sanders12a}. However, even with Gaia, stream data in the future
will typically have large uncertainties compared to the intrinsic
dispersion of the stream and in particular the line-of-sight
velocities of the faintest stream members will not be observed for
many stars. Additionally, streams are superimposed on a non-negligible
background of field stars and background contamination cannot be
easily taken into account in any of the current stream-fitting
methods.

In this paper, I present a new method for modeling cold tidal streams
that fundamentally lives in frequency--angle space, because after
specifying the time at which a star was stripped, the stream
distribution function is essentially one-dimensional in this
space. While many of the results of this paper are more generally
valid, the fiducial stream model in this paper consists of a
three-dimensional (close to) Gaussian distribution of frequencies in
the stream, a three-dimensional Gaussian distribution of initial angle
offsets between stream members and the progenitor, and a uniform
distribution of stripping times; the frequency and angle distributions
are independent of stripping time. While this model oversimplifies the
real stripping process, it is argued that at least for constraining
the gravitational potential with stream data, this simple model is
likely unbiased. For detailed mock stream data, more elaborate models
of the stripping process as a function of time might be necessary and
I discuss how these could easily be incorporated into the proposed
framework.

To evaluate the model in position--velocity coordinates, the
transformation to and from action--angle coordinates is computed using
a novel method for calculating action--angle coordinates presented in
Appendix~\ref{sec:aa} and this transformation is linearized at a small
number of points along the progenitor orbit that span the width of the
stream. For cold streams, the separation between the stream track and
the progenitor orbit as well as the internal differences within the
stream are small, such that this linear approximation is accurate
enough. The linear approximation allows for fast evaluation of the
average stream position in position--velocity coordinates and
efficient mock data generation. For a uniform distribution of
stripping times, an observed stream member's phase--space probability
distribution function (PDF) can be analytically marginalized over
stripping time, such that using the linear approximation any
evaluation of the PDF is extremely rapid. This clears the way for
proper probabilistic inference of the Milky Way halo's gravitational
potential and progenitor properties using stream data for individual
stars.

This paper is outlined as follows. In \sectionname~\ref{sec:dynamics}
I briefly summarize the dynamics of tidal streams in terms of
action--angle coordinates. A generative model of a tidal stream in
frequency--angle coordinates is given in
\sectionname~\ref{sec:generative}. \sectionname~\ref{sec:track}
describes how to compute the stream's properties as a function of
angle along the stream, both in action--angle coordinates and in
poition--velocity coordinates. \sectionname~\ref{sec:mock} discusses
how to generate mock stream data using the generative model of a tidal
stream and in \sectionname~\ref{sec:pdf} I show how to calculate the
stream PDF for individual stream members as well as its
marginalization and convolution over missing and noisy directions. A
discussion and outlook is presented in
\sectionname~\ref{sec:discussion} and I conclude in
\sectionname~\ref{sec:conclusion}. In Appendix
\sectionname~\ref{sec:aa} a novel method for calculating actions,
frequencies, and angles for any orbit in any static gravitational
potential is presented.

@@

The dynamics of tidal streams both in real (position--velocity) and
action--angle space has been described by various previous authors
\citep[\eg,][]{Helmi99a,Tremaine99a,SomeJohnstonPaper,Sanders13b}. I
provide here a brief description of the dynamics of tidal streams that
is relevant for what follows.

@@

Thus, in action--angle coordinates, the dynamics of tidal-stream
formation is simplified to linear growth of small initial angle
differences due to small action/frequency differences. The angle
direction $\unitvec{\veco} \equiv \Delta \veco / | \Delta \veco|$
along which a star moves away from the progenitor is related to the
initial action offset by $\Delta \veco \approx (\partial^2 H
/ \partial \vecj \partial \vecj) \Delta \vecj$, in the limit of small
action offsets, as $\veco = \partial H / \partial \vecj$. For an
approximately one-dimensional stream to form, the Hessian matrix
$(\partial^2 H / \partial \vecj \partial \vecj)\large|_{\vecj^p}$ has
to be dominated by a single large eigenvalue; in this case the angle
difference of stars stripped from the stream at any time will fall
along approximately the same direction $\unitvec\veco$, and the stream
will essentially be one dimensional. In detail, because the
distribution of initial offsets $\Delta \vecj$ is not isotropic, the
direction $\unitvec \veco$ is not just determined by the potential
(through $H$) and the progenitor orbit, but also by the distribution
of $\Delta \vecj$, which is set through a combination of the internal
dynamics of the progenitor and the stripping process (see discussion
in \citealt{Sanders13b}). For dynamical modeling we can be agnostic
about the relative importance of all of these contributors to
$\unitvec \veco$ by flexible modeling of the distribution of $\Delta
\vecj$.

@@

The probability in \eqnname~(\ref{eq:aadfsimple}) incorporates and
elucidates commonly used methods for fitting tidal streams. On the one
hand are approaches that search to minimize the spread in energy,
integrals of the motion, or actions (\eg,
\citealt{Binney08a,Penarrubia12a}, Sanderson \etal, 2013, in
preparation). These approaches ignore the correlations between actions
and angles in the stream (the exponential in
\eqnname~(\ref{eq:aadfsimple}), which describes this correlation
through the frequencies) and only use $f_{\vecj}(\Delta
\vecj_i|\paramsdiff)$; finding the gravitational potential that
minimizes the spread in actions optimizes $f_{\vecj}(\Delta
\vecj_i|\paramsdiff)$. Recently, a method was proposed that only uses
the correlation between the actions and the angles, expressed by the
exponential in \eqnname~(\ref{eq:aadfsimple}) \citep{Sanders13b}; this
method does not use the fact that the spread in actions in a tidal
stream is small (expressed by $f_{\vecj}(\Delta
\vecj_i|\paramsdiff)$).

@@

\bibitem[{{Binney} \& {Tremaine}(2008)}]{binneytremaine}
  Binney,~J. \& Tremaine,~S. 2008, Galactic Dynamics: Second Edition
\bibitem[Binney(2008)]{Binney08a}
  Binney,~J. 2008, \mnras, 386, L47
\bibitem[Carlberg(2009)]{Carlberg09a}
  Carlberg,~R.~G. 2009, \apj, 705, L223
\bibitem[Carlberg(2012)]{Carlberg12a}
Carlberg,~R.~G. 2012, \apj, 748, 20
\bibitem[Eyre \& Binney(2011)]{Eyre11a}
  Eyre,~A. \& Binney,~J. 2011, \mnras, 413, 1852
\bibitem[Helmi \& White(1999)]{Helmi99a}
  Helmi,~A. \& White,~S.~D.~M. 1999, \mnras, 307, 495
\bibitem[Johnston \etal(1999)]{Johnston99a}
  Johnston,~K.~V., Zhao,~H., Spergel,~D.~N., \& Hernquist,~L. 1999, \apj, 512, L109
\bibitem[Koposov \etal(2010)]{Koposov10a}
  Koposov,~S.~E., Rix,~H.-W., \& Hogg,~D.~W. 2010, \apj, 712, 260
\bibitem[Ngan \& Carlberg(2014)]{Ngan14a}
  Ngan,~W.~H.~W. \& Carlberg,~R.~G. 2014, \apj, submitted
\bibitem[Pe{\~n}arrubia \etal(2012)]{Penarrubia12a}
  Pe{\~n}arrubia, J., Koposov,~S.~E., \& Walker,~M.~G. 2012, \apj, 760, 2
\bibitem[Price-Whelan \& Johnston(2013)]{PriceWhelan13a}
  Price-Whelan,~A.~M. \& Johnston,~K.~V.~2013, \apj, 778, L12
\bibitem[Sanders(2012)]{Sanders12a}
  Sanders,~J. 2012, \mnras, 426, 128
\bibitem[Sanders \& Binney(2013a)]{Sanders13a}
  Sanders,~J. \& Binney,~J. 2013a, \mnras, 433, 1813
\bibitem[Sanders \& Binney(2013b)]{Sanders13b}
  Sanders,~J. \& Binney,~J. 2013b, \mnras, 433, 1826
\bibitem[Tremaine(1999)]{Tremaine99a}
  Tremaine,~S. 1999, \mnras, 307, 877
\end{thebibliography}
